Lessons learned from bugs in models of human history

### 2020-08-05 paper.tex
remove shading around code blocks

\documentclass{article}
\usepackage[super,comma]{natbib}
\usepackage{fullpage}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{minted}
\usepackage{tcolorbox}
\usepackage{etoolbox}
\BeforeBeginEnvironment{minted}{\begin{tcolorbox}}%
\AfterEndEnvironment{minted}{\end{tcolorbox}}%
\usepackage{url}

@@

The \msprime\ coalescent
simulator~\citep{kelleher2016efficient,nelson2020accounting,kelleher2020coalescent}
is now widely used in genetics studies.
Much of its appeal is the large increase in efficiency over the classical
\ms\ program~\citep{hudson2002generating} which makes it feasible to simulate large
samples of whole chromosomes for the first time. Another distinct advantage
of \msprime\ is its Python application programming interface (API), which greatly
increases the flexibility and ease of use over the standard approach of
text-based command line interfaces.
In particular, programs like \ms\ require users to specify cryptic command
line options to describe demographic models.
For example, the~\citet{gutenkunst2009inferring} demographic model
(that is the subject of this note), as written
in \ms\ syntax, is
\begin{minted}[]{tcsh}
-n 1 1.682020 -n 2 3.736830 -n 3 7.292050
-eg 0 2 116.010723 -eg 0 3 160.246047
-ma x 0.881098 0.561966 0.881098 x 2.797460 0.561966 2.797460 x
-ej 0.028985 3 2 -en 0.028985 2 0.287184
-ema 0.028985 3 x 7.293140 x 7.293140 x x x x x
-ej 0.197963 2 1 -en 0.303501 1 1
\end{minted}
\noindent This model is relatively simple,
and models with many more populations and parameters are increasingly
common. Descriptions of such models in \ms\ syntax are not easy to comprehend.
The Python interface for \msprime, by contrast, allows the user to
state models in a more human-readable and programmatic manner,
and has many advantages over \ms's command line interface.

@@

\begin{minted}[linenos,highlightlines={15},numbersep=5pt]{python}
demographic_events = [
    # CEU and CHB merge into B with rate changes at T_EU_AS
    msprime.MassMigration(
        time=T_EU_AS, source=2, destination=1, proportion=1.0),
    msprime.MigrationRateChange(time=T_EU_AS, rate=0),
    msprime.MigrationRateChange(
        time=T_EU_AS, rate=m_AF_B, matrix_index=(0, 1)),
    msprime.MigrationRateChange(
        time=T_EU_AS, rate=m_AF_B, matrix_index=(1, 0)),
    msprime.PopulationParametersChange(
        time=T_EU_AS, initial_size=N_B, growth_rate=0, population_id=1),
    # Population B merges into YRI at T_B
    msprime.MassMigration(
        time=T_B, source=1, destination=0, proportion=1.0),
    msprime.MigrationRateChange(time=T_B, rate=0), # NB This line was missing!!!
    # Size changes to N_A at T_AF
    msprime.PopulationParametersChange(
        time=T_AF, initial_size=N_A, population_id=0)
]
\end{minted}
Note that the migration rate change on 15 was missing in the incorrect model.
The release of \msprime\ version~1.0 will introduce a \texttt{PopulationSplit} event,
which more intuitively links the movement of lineages with appropriate changes in
migration rates at the time of the split.

@@

From a software perspective, this error was easy to make
and could have been prevented by better API design.
The original \msprime\ API requires the user to pass
three separate parameters to specify a demographic model. For example,
\begin{minted}[highlightlines={8}]{python}
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  demographic_events=demographic_events,
  migration_matrix=migration_matrix)
dbg.print_history()
ts = msprime.simulate(
  population_configurations=population_configurations,
  demographic_events=demographic_events, # Missing in Martin et al. 2017
  migration_matrix=migration_matrix)
\end{minted}
Note that the same three parameters must be passed to both the debugger
and the simulate function.
To help prevent such errors,
\msprime~ver.~1.0 will introduce a Demography class. The above
snippet would then be rewritten as
\begin{minted}[]{python}
demography = msprime.Demography(
  population_configurations=population_configurations,
  demographic_events=demographic_events,
  migration_matrix=migration_matrix)
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(demography=demography)
\end{minted}
This simple change to the interface makes it much less likely
that different models are passed to the debugger and simulator,
reducing the potential for error.

@@

\section*{Acknowledgements}
We would like to thank Alicia Martin for comments on the manuscript
and for providing the simulation data from~\cite{martin2017human}
used in Fig.~\ref{fig:prs}. We also like to thank
Chris Gignoux, Ilan Gronau and John Novembre for helpful comments on the manuscript.
JK would like to apologise
to those affected by the error in the example Out-of-Africa model and
by the poor design decisions in \msprime's demography API.

>>>>

\documentclass{article}
\usepackage[super,comma]{natbib}
\usepackage{fullpage}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{minted}
\usepackage{tcolorbox}
\usepackage{etoolbox}
\BeforeBeginEnvironment{minted}{\begin{tcolorbox}}%
\AfterEndEnvironment{minted}{\end{tcolorbox}}%
\usepackage{url}
\usepackage{fancyvrb}

@@

The \msprime\ coalescent
simulator~\citep{kelleher2016efficient,nelson2020accounting,kelleher2020coalescent}
is now widely used in genetics studies.
Much of its appeal is the large increase in efficiency over the classical
\ms\ program~\citep{hudson2002generating} which makes it feasible to simulate large
samples of whole chromosomes for the first time. Another distinct advantage
of \msprime\ is its Python application programming interface (API), which greatly
increases the flexibility and ease of use over the standard approach of
text-based command line interfaces.
In particular, programs like \ms\ require users to specify cryptic command
line options to describe demographic models.
For example, the~\citet{gutenkunst2009inferring} demographic model
(that is the subject of this note), as written
in \ms\ syntax, is
%\begin{minted}[]{tcsh}
\begin{indent}
\begin{Verbatim}[xleftmargin=.5in]
-n 1 1.682020 -n 2 3.736830 -n 3 7.292050
-eg 0 2 116.010723 -eg 0 3 160.246047
-ma x 0.881098 0.561966 0.881098 x 2.797460 0.561966 2.797460 x
-ej 0.028985 3 2 -en 0.028985 2 0.287184
-ema 0.028985 3 x 7.293140 x 7.293140 x x x x x
-ej 0.197963 2 1 -en 0.303501 1 1
\end{Verbatim}
\end{indent}
%\end{minted}
\noindent This model is relatively simple,
and models with many more populations and parameters are increasingly
common. Descriptions of such models in \ms\ syntax are not easy to comprehend.
The Python interface for \msprime, by contrast, allows the user to
state models in a more human-readable and programmatic manner,
and has many advantages over \ms's command line interface.

@@

%\begin{minted}[linenos,highlightlines={15},numbersep=5pt]{python}
\begin{Verbatim}[xleftmargin=.5in]
demographic_events = [
    # CEU and CHB merge into B with rate changes at T_EU_AS
    msprime.MassMigration(
        time=T_EU_AS, source=2, destination=1, proportion=1.0),
    msprime.MigrationRateChange(time=T_EU_AS, rate=0),
    msprime.MigrationRateChange(
        time=T_EU_AS, rate=m_AF_B, matrix_index=(0, 1)),
    msprime.MigrationRateChange(
        time=T_EU_AS, rate=m_AF_B, matrix_index=(1, 0)),
    msprime.PopulationParametersChange(
        time=T_EU_AS, initial_size=N_B, growth_rate=0, population_id=1),
    # Population B merges into YRI at T_B
    msprime.MassMigration(
        time=T_B, source=1, destination=0, proportion=1.0),
    msprime.MigrationRateChange(time=T_B, rate=0), # NB This line was missing!!!
    # Size changes to N_A at T_AF
    msprime.PopulationParametersChange(
        time=T_AF, initial_size=N_A, population_id=0)
]
\end{Verbatim}
%\end{minted}
Note that the migration rate change on 15 was missing in the incorrect model.
The release of \msprime\ version~1.0 will introduce a \texttt{PopulationSplit} event,
which more intuitively links the movement of lineages with appropriate changes in
migration rates at the time of the split.

@@

From a software perspective, this error was easy to make
and could have been prevented by better API design.
The original \msprime\ API requires the user to pass
three separate parameters to specify a demographic model. For example,
%\begin{minted}[highlightlines={8}]{python}
\begin{Verbatim}[xleftmargin=.5in]
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  demographic_events=demographic_events,
  migration_matrix=migration_matrix)
dbg.print_history()
ts = msprime.simulate(
  population_configurations=population_configurations,
  demographic_events=demographic_events, # Missing in Martin et al. 2017
  migration_matrix=migration_matrix)
\end{Verbatim}
%\end{minted}
Note that the same three parameters must be passed to both the debugger
and the simulate function.
To help prevent such errors,
\msprime~ver.~1.0 will introduce a Demography class. The above
snippet would then be rewritten as
%\begin{minted}[]{python}
\begin{Verbatim}[xleftmargin=.5in]
demography = msprime.Demography(
  population_configurations=population_configurations,
  demographic_events=demographic_events,
  migration_matrix=migration_matrix)
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(demography=demography)
\end{Verbatim}
%\end{minted}
This simple change to the interface makes it much less likely
that different models are passed to the debugger and simulator,
reducing the potential for error.

@@

\section*{Acknowledgements}
We would like to thank Alicia Martin for comments on the manuscript
and for providing the simulation data from their study reproduced
in Fig.~\ref{fig:prs}. We also like to thank
Chris Gignoux, Ilan Gronau and John Novembre for helpful comments on the manuscript.
JK would like to apologise
to those affected by the error in the example Out-of-Africa model and
by the poor design decisions in \msprime's demography API.
